---
title: "checktor in Continuous Integration"
vignette: >
  %\VignetteIndexEntry{checktor in Continuous Integration}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(checktor)
```

## Why bother

`R CMD check` answers one question very well: *does this package build and run?*
It does not answer the question that actually decides your submission: *will a
CRAN volunteer, reading by hand, send it back?* Those are different questions,
and the gap between them is where afternoons disappear: a title that isn't in
title case, a `\value{}` tag you forgot, a stray `T` standing in for `TRUE`.

The fix is not vigilance. Vigilance does not survive contact with a deadline.
The fix is to let a machine remember the rules so you don't have to, and the
natural home for a machine that never forgets is your continuous integration
(CI) pipeline. Run `checktor` there and every push gets a second opinion for
free.

## The one function CI needs

Everything in this article rests on a single function. `checkup()` runs the
full diagnosis and collapses it to one verdict: `TRUE` if the package is clean,
`FALSE` if anything wants attention.

```{r}
#| label: scenario
# A throwaway package that (deliberately) uses T/F instead of TRUE/FALSE
bad_pkg <- example_diagnose_scenario("code_examples/tf_usage_bad.R")

checkup(bad_pkg)
```

A healthy package returns `TRUE`. That is the entire contract, and it is all a
build needs to decide whether to go green or red.

## A GitHub Actions job

`checkup()` is designed to be the last word in a shell one-liner. Drop this file
in at `.github/workflows/checktor.yaml` and you have an extra-CRAN gate on every
push and pull request:

```yaml
name: checktor

on:
  push:
    branches: [main, master]
  pull_request:

jobs:
  checktor:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: coatless-rpkg/checktor

      - name: Run extra-CRAN checks
        run: if (!checktor::checkup()) quit(status = 1)
        shell: Rscript {0}
```

The job checks out your package, installs `checktor`, and exits non-zero the
moment `checkup()` disagrees with you. (Once `checktor` is on CRAN, swap the
`extra-packages:` line for `any::checktor`.)

## Failing loudly, not silently

A red X that says only "exit code 1" is a riddle, not a report. When a build
fails, you want the diagnosis in the log, not a scavenger hunt. Run the full
`checktor()` once and print the report before you quit:

```r
results <- checktor::checktor()

if (results$metadata$total_issues > 0) {
  cat(checktor::health_report(results), sep = "\n")
  quit(status = 1)
}
```

`health_report()` returns the findings as Markdown, so the failing log reads
like a chart at the foot of a hospital bed: what is wrong, and where. Point its
`file =` argument at a path and you can just as easily upload the report as a
build artifact for the squeamish who prefer not to read CI logs.

## Tuning the examination

CI is chatty by nature, so quiet the doctor down. `configure_doctor()` sets the
verbosity and progress defaults once, at the top of a script, and every later
call inherits them:

```r
checktor::configure_doctor(verbose_default = FALSE, progress_default = FALSE)
```

You can also run a single category when that is all you care about, since each
`diagnose_*_issues()` function stands on its own:

```r
# Gate only on DESCRIPTION-field problems
desc <- checktor::diagnose_description_issues(".")
```

## Closer to the keyboard

CI is the backstop, not the first line. If you would rather hear about a problem
before it reaches a pull request, the same one-liner works as a local Git
pre-commit hook:

```bash
# .git/hooks/pre-commit  (make it executable: chmod +x)
#!/usr/bin/env bash
Rscript -e 'if (!checktor::checkup()) quit(status = 1)'
```

Now a commit that would have embarrassed you on CRAN never leaves your laptop.

## The takeaway

`R CMD check` is your build's physician; `checktor` is the specialist it refers
you to before the operation. Wire `checkup()` into CI once and the rules stop
living in your head, where they were never safe anyway. The best submission is
the one where the reviewer finds nothing to say — and the surest way there is to
have already said it to yourself.

## See also

- [Getting Started with checktor](getting-started-with-checktor.html) — the
  guided tour of the diagnostics.
- [Writing Your Own Checks](writing-checks.html) — add project-specific checks
  against the parsed AST.
