---
title: "checktor in Continuous Integration"
vignette: >
  %\VignetteIndexEntry{checktor in Continuous Integration}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(checktor)
```

<style>
.ct-fig{display:block;margin:1.5rem auto;max-width:100%;height:auto}
.ct-dark{display:none}
@media (prefers-color-scheme:dark){.ct-light{display:none}.ct-dark{display:block}}
[data-bs-theme=light] .ct-light{display:block}
[data-bs-theme=light] .ct-dark{display:none}
[data-bs-theme=dark] .ct-light{display:none}
[data-bs-theme=dark] .ct-dark{display:block}
</style>

You already run `checktor` on your package. The trouble with running it by hand
is that you do it when you remember to, which is not the same as every time it
matters. A pipeline has no such lapses, running the check on every push, for
every contributor, whether or not anyone remembered to.

Moving the check into continuous integration (CI) turns a good habit into an
unconditional one, in about as many lines of YAML as it takes to describe.

## The one function CI needs

Everything in this article rests on a single function. `checkup()` runs the
full diagnosis and collapses it to one verdict: `TRUE` if the package is clean,
`FALSE` if anything wants attention.

```{r}
#| label: scenario
# A throwaway package that deliberately uses T/F instead of TRUE/FALSE
bad_pkg <- example_diagnose_scenario("code_examples/tf_usage_bad.R",
                                     show_content = FALSE)

checkup(bad_pkg)
```

A healthy package returns `TRUE`. That is the entire contract, and it is all a
build needs to decide whether to go green or red.

## A GitHub Actions job

`checkup()` is designed to be the last word in a shell one-liner. Drop this file
in at `.github/workflows/checktor.yaml` and you have an extra-CRAN gate on every
push and pull request:

```yaml
name: checktor

on:
  push:
    branches: [main, master]
  pull_request:

jobs:
  checktor:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v7

      - uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: coatless-rpkg/checktor

      - name: Run extra-CRAN checks
        run: if (!checktor::checkup()) quit(status = 1)
        shell: Rscript {0}
```

The job checks out your package, installs `checktor`, and exits non-zero the
moment `checkup()` disagrees with you.

> **Once `checktor` is on CRAN**, swap the `extra-packages:` line for
> `any::checktor`, which needs no GitHub remote.

## Failing loudly, not silently

A red X that says only "exit code 1" is a riddle, not a report. When a build
fails, you want the diagnosis in the log, not a scavenger hunt. Run the full
`checktor()` once and print the report before you quit:

```r
results <- checktor::checktor()

if (!checktor::is_healthy(results)) {
  writeLines(checktor::health_report(results))  # the diagnosis, in the log
  quit(status = 1)                              # then fail the build
}
```

`health_report()` returns the findings as Markdown, so the failing log reads
like a chart at the foot of a hospital bed, showing what is wrong and where.
Point its `file =` argument at a path and you can just as easily upload the
report as a build artifact for the squeamish who prefer not to read CI logs.

## Tuning the examination

CI is chatty by nature, so quiet the doctor down. `configure_doctor()` sets the
verbosity and progress defaults once, at the top of a script, and every later
call inherits them:

```r
checktor::configure_doctor(verbose_default = FALSE, progress_default = FALSE)
```

You can also run a single category when that is all you care about, since each
`diagnose_*_issues()` function stands on its own:

```r
# Gate only on DESCRIPTION-field problems
desc <- checktor::diagnose_description_issues(".")
```

## Closer to the keyboard

CI is the backstop, not the first line. If you would rather hear about a problem
before it reaches a pull request, the same one-liner works as a local Git
pre-commit hook:

```bash
# .git/hooks/pre-commit  (make it executable: chmod +x)
#!/usr/bin/env bash
Rscript -e 'if (!checktor::checkup()) quit(status = 1)'
```

Now a commit that would have embarrassed you on CRAN never leaves your laptop.

It is the same call in both places, and the only thing that changes is how long
you wait to hear about it.

![The same `checkup()` call at three latencies. In a `.git/hooks/pre-commit` hook on your laptop it answers in seconds. In GitHub Actions on the pull request it answers in minutes. If you skip both, the same finding reaches you from a CRAN reviewer weeks later.](figures/ci-gates-light.svg){.ct-fig .ct-light}
![](figures/ci-gates-dark.svg){.ct-fig .ct-dark aria-hidden="true"}

## The takeaway

Wire `checkup()` into CI once and the rules stop living in your head, where they
were never safe anyway. The best submission is the one where the reviewer finds
nothing to say, and the surest way there is to have already said it to yourself,
automatically, on every push.

## See also

- [Getting Started with checktor](getting-started-with-checktor.html): the
  guided tour of the diagnostics.
- [Writing Your Own Checks](writing-checks.html): add project-specific checks
  against the parsed syntax tree.
